---
title: "`MADloy`: Detecting mosaic loss of Y events on SNP and NGS data"
subtitle: "Marcos Lopez-Sanchez, Juan R. Gonzalez"
author: |
  | Institute for Global Health (ISGlobal), Barcelona, Spain
  | Bioinformatics Research Group in Epidemiolgy (BRGE)
  | (<http://www.creal.cat/brge.htm>)
date: "`r format(Sys.Date(), '%d %B %Y')`"
package: "`r paste('MADloy', packageVersion('MADloy'))`"
output:    
  BiocStyle::html_document:
    number_sections: true
    toc: yes
    fig_caption: yes
    fig_height: 3
    fig_width: 4
vignette: >
  %\VignetteIndexEntry{`MADloy`: Detecting mosaic loss of Y events (LOY) on SNP and NGS data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
# Introduction

`MADloy` is a package to detect mosaic loss of chromosome Y (LOY) events from NGS and SNP data. It uses a bivariate mixture model which uses values in a target region of chromosome Y and in two reference chromosomes across several samples in order to detect those samples with altered values (LOY or GAIN) from the main population. The measures compared are different depending the type of data analyzed. Depth coverage is used for NGS data (exome sequencing data) and log R ratio (LRR) for SNP array data.

By default, the region used to detect LOY is a region in chromosome Y region that avoids the PAR1 and PAR2 regions shared with chromosome X (mLRR-Y). In order to take into consideration the variability of coverage or LRR in a given sample, the method clusters individuals into LOY, normal and GAIN categories by using a bivariate vector having data at mLRR-Y region and whole chromosome 22. Nonetheless, the reference region can be set manually. The package summarizes the coverage (median, mean and trimmed-mean) in all sequenced positions for the analyzed region when having NGS data, and the median LRR in the probes contained in the SNP array. These bivariate observations are computed for each individual and a bivariate mixture model is fitted in order to: 1st) decide how many cluster are in the data (2: LOY and Normal or 3: LOY, Normal and GAIN) and 2nd) call individuals.


The method to detect the altered values is different for each type of data. **Marcos, esto no se a que te refieres** **Me refiero a que el método para detectar los valores alterados es diferente para los datos de secuencia y para los datos de array**


You can install `MADloy` from Github by typing 

```{r eval=FALSE}
devtools::install_github("isglobal-brge/MADloy")
```

```{r, message = FALSE}
library(MADloy)
```

# Loading data and summarizing data

`MADloy` contains a different function to pre-process your data depending on the technology. `madloy` processes SNP array data, while `madseqloy` deals with Exome Sequencing data. Next, we illustrate how to preform  this step in both situations.

## Processing SNP data with `madloy`

Detecting LOY events in SNP arrays requires to get summarized data from LRR files. If your data has been obtained from Illumina arrays, LRR can be obtained from the reports generated with BeadStudio tool (see for instance `MAD` package documentation [here](http://www.creal.cat/media/upload/arxius/jr/GADA/mad_vignette.pdf)), or other R packages as [`crlmm`](http://bioconductor.org/packages/release/bioc/html/crlmm.html) or [`beadarraySNP`](https://www.bioconductor.org/packages/release/bioc/html/beadarraySNP.html). If your data has been generated from any Affymetrix chip (e.g. GenomeWide, Axiom or CytoScan) you can use the either Affymetrix Power Tools (APT) or `affy2sv` package **Marcos un link**. Files having LRR of each sample must be included in a folder called "rawData" **Marcos, si rawDataPath ya es el rawData, no tiene sentido que digamos que tiene que tener una carpeta que se llame asi** **Era por mantener a nomenclatura del `MAD`**. The function requires a path indicanting were those files are located. `MADloy` package contains some files having the LRR of 3 samples. The folder can be get by writting:


```{r, MADloy_1}
rawDataPath <- system.file("extdata", "rawData", package = "MADloy")
rawDataPath
dir(rawDataPath)
```
Then, data can be summarized by executing:
```{r, MADloy_1_b}
res <- madloy(rawDataPath)
res
```
Otherwise, a path to the files can be used instead of the folder path to specify which samples must be analyzed
```{r, MADloy_2}
files <- list.files(rawDataPath, full.names=T)
files
res <- madloy(files[1])
res
```
## Visualization
Results of this initial processing can be visualized by plotting the {\tt MADloy} object. Let's us image that we have analyzed 89 CEU HapMap 3 male samples. This preprocess data can be loaded in to your ``R`` session from the `MADloyDataAll` object that is included in the package. **Take into account the gender of the samples analyzed, because female samples are going to have large negative LRR values in chromosome Y**.

```{r, MADloy_3}
data(MADloyDataAll)
MADloyDataAll
```
```{r, fig.show='hold', fig.width=8, fig.height=6, fig.cap = "Plot of a MADloy object with all HapMap 3 samples analyzed"}
plot(MADloyDataAll)
```

As warned before, due to the lack of chromosome Y in female samples, the plot displays a large number of samples with the median difference between -2 and -6. Those are validated female individuals and it is not possible to have LOY events. In order to focus the attention in the male individuals, 23 male samples of those with Exome-seq data have been selected and processed in the `MADloyDataMale` object.

```{r, MADloy_4}
data(MADloyDataMale)
MADloyDataMale
```

```{r, fig.show='hold', fig.width=8, fig.height=6, fig.cap = "Plot of a MADloy object with 23 HapMap 3 male samples analyzed"}
plot(MADloyDataMale, ylim=c(-0.3, 0.3))
```

## Processing NGS Exome-seq data with `madseqloy`

To evaluate the LOY events for NGS data, the procedure is almost the same used for SNP data, but having into account that is necessary to input the file path to the targets file when analyzing exome-seq data. The next code downloads and analyzes a Exome-seq data sample from HapMap 3 project. This code is not executed due to the large size of Exome-seq data files, but you can try to run it. It is required that .bam files are indexed, in other words, that the .bai index file is in the same folder and with the same filename in order to proceed (b.e., test.bam and test.bai). If you don't have an index file, you can create with the function `indexBam` from `Rsamtools` package considering that the .bam file is sorted. Otherwise, you will need to sort it beforehand with the function `sortBam` from the same package.

```{r, Download_Hapmap, eval=FALSE}
download.file(url = "ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/data/HG12716/exome_alignment/HG00236.mapped.ILLUMINA.bwa.GBR.exome.20121211.bam",
              destfile = "/tmp/HG00236.mapped.ILLUMINA.bwa.GBR.exome.20121211.bam.bai")
download.file(url = "ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase3/data/HG00236/exome_alignment/HG00236.mapped.ILLUMINA.bwa.GBR.exome.20121211.bam.bai",
              destfile = "/tmp/HG00236.mapped.ILLUMINA.bwa.GBR.exome.20121211.bam.bai")
```

The targets file for HapMap analyzed samples is "SureSelect Human All Exon V2". It must be downloaded from Agilent webpage [https://earray.chem.agilent.com/suredesign/search.htm](https://earray.chem.agilent.com/suredesign/search.htm) with the Design ID S0293689, but registration is required. To follow the tutorial, it is considered that the targets file is decompressed and inside the /tmp folder.

```{r, MADseqLOY_1, eval = FALSE}
res <- madseqloy(files = "/tmp/HG00236.mapped.ILLUMINA.bwa.GBR.exome.20121211.bam", exomeTargets = "/tmp/20130108.exome.targets.bed")
```
As with the `madloy` function, you can use the folder path instead of each file path to select which samples should be analyzed. The function search for all .bam files in the inputted path recursively, so you can put a path that stores a large number of .bam files, but you still are required to input the exomeTargets path.

```{r, MADseqLOY_2, eval = FALSE}
res <- madseqloy(files = "/tmp", exomeTargets = "/tmp/S0293689/S0293689_Padded.bed")
```
The samples analyzed can be visualized one by one from the {\tt MADseqLOY} object using the generic `plot` function, but is recommended to have a minimum of 4GB of RAM memory to draw the plot. Otherwise, it could crash or take a large amount of time. The next code and image summarizes that. In order to plot a sample, it is required to input the filename of the sample to be plot. The {\tt MADseqLOY} object stores the files in a vector in test$par$files where test is the name of the {\tt MADseqLOY} object.

```{r, MADseqLOY_3, eval = FALSE}
plot(res, sample = res$par$files[1])
```

![Plot of a female HapMap sample in a MADseqLOY object](`r file.path(system.file("extdata", package = "MADloy"), "Rplot001.png")`)

# Detecting LOY events

Once samples are processed, you will have a {\tt MADloy} or {\tt MADseqLOY} object. The next step is detect which of the processed samples have a Loss of Y event using the getLOY function.

```{r, getLOY}


```

Results can be visualized with the plot function as follows in this example:

```{r, fig.show='hold', fig.cap = "Plot of a LOY object"}


```